Mendelian randomization for causal inference in Genome-wide association studies

### 2022-02-19 mr-review.Rmd
deleted multiple outcomes section

Many approaches to MR can be classified into two categories: two-stage methods and 
likelihood-based methods. A two-stage statistical model involves two regression models [@burgess_review_2017;@sawa_exact_1969]. For a continuous 
outcome, one may perform "two-stage least squares", which involves two linear regressions. 
First, the exposure variable is regressed on the instrument (*i.e.*, the SNP genotypes) (Equation \@ref(eq:lm1)). The resulting 
fitted values for the exposure variable are the independent variable in the second linear regression, 
where the outcome variable is the dependent variable. The causal effect estimate, then, is the 
coefficient $\hat \beta_1$ obtained from the second regression analysis (Equation \@ref(eq:lm2)).

@@

where the $\epsilon_i$ are assumed to be independent and identically distributed normal random variables with mean zero.

@@

where the $\tau_i$ are assumed to be independent and identically distributed  normal random variables with mean zero and $\hat x_i$ denotes the fitted value for $x_i$ obtained from Equation \@ref(eq:lm1).

@@

Likelihood-based MR methods, unlike two-stage methods, provide maximum likelihood 
estimates with their many desirable properties [@fisher1925theory]. Limited information maximum likelihood from econometrics 
is the earliest approach for likelihood-based inference in MR [@davidson1993estimation]. CoMM and 
PMR-Egger are two recent likelihood-based methods [@yang_comm_2019;@yuan_testing_2020]. We discuss 
PMR-Egger below in the context of methods that account for horizontal pleiotropy. Immediately below, we 
discuss CoMM.

@@

Now that we've discussed several MR methods, it is worthwhile to consider ways in which the standard MR assumptions may be violated. 
Let's first consider the exclusion restriction assumption (assumption 2 above). It states that the genetic marker 
must affect the outcome only through the exposure. The causal diagram in Figure \@ref(fig:dag1) illustrates 
a scenario that violates this assumption when the variable X is used as the exposure [@vanderweele_methodological_2014]. Bias in causal effect estimation would result. However, 
if variable W is measured and treated as the exposure, then an unbiased estimate of causal effect is 
possible, since W captures the two pathways, W -> Y and W -> X -> Y, through which the genetic marker G 
affects the outcome Y [@vanderweele_methodological_2014;@textor2011dagitty;@textor2016robust].

@@

Sensitivity analysis is recommended and widely used in MR studies because of the inability to verify 
the three MR assumptions with observational data. The goal of a sensitivity analysis is to gain insight 
into how the results might differ if the assumptions are violated.

Because not all confounding 
variables are known or measured, assumptions 2 and 3 are not fully verifiable 
in MR studies. Recognition of this fact has inspired the development of 
sensitivity analysis tools. While we can't assess whether a SNP instrument is 
associated with every confounder of the exposure-outcome association, it is 
possible to examine the associations of the SNP instrument with the measured 
covariates. While absence of such associations doesn't guarantee satisfaction 
of the assumption, presence of SNP-covariate associations must be investigated
carefully, as they may constitute assumption violations [@burgess_sensitivity_2017].

@@

Recent research on the omnigenic hypothesis, which posits 
that every SNP's effect on a trait is nonzero, has informed
MR methods development [@Boyle2017]. @wang_mendelian_nodate 
recognized the limitations of previous MR methods in the 
context of the omnigenic hypothesis. In response, @wang_mendelian_nodate 
developed a MR method that explicitly uses all genome-wide SNPs as instruments. 
Using GWAS summary statistics as inputs, their method relies on a composite likelihood framework for scalable computation and allows for horizontal pleiotropy. @wang_mendelian_nodate used extensive 
simulations, including those with model misspecifications, to characterize their method's statistical power and robustness. Finally, they applied the new method to 
identify multiple complex traits that affect coronary artery disease and asthma. These causal relationships highlight the important roles of plasma lipids, blood pressure, and the immune system in CAD susceptibility and those of obesity and the immune system in asthma development.

## Multiple outcomes in MR

Investigators have found that combining information from multiple related 
traits can lead to richer MR results [@liu_multi-trait_2021].

One approach to combining information across traits is to extend the 
univariate MR statistical methods to their multivariate analogs.

>>>>

Many approaches to MR can be classified into two categories: two-stage methods and 
likelihood-based methods. A two-stage statistical model involves two regression models [@burgess_review_2017;@sawa_exact_1969]. For a continuous 
outcome, one may perform "two-stage least squares", which involves two linear regressions. 
First, the exposure variable is regressed on the instrument (*i.e.*, the SNP genotypes) (Equation \@ref(eq:lm1)). We denote each subject's exposure variable value as $x_i$ in Equation \@ref(eq:lm1), while $\alpha_0$ denotes an intercept term and $\alpha_1$ is the slope. $z_i$ denotes the SNP instrument genotype for subject $i$, while $\epsilon_i$, the error term, is assumed to be independent among subjects and normally distributed with a shared common variance and mean zero.

The resulting 
fitted values for the exposure variable from Equation \@ref(eq:lm1) are the independent variable in the second linear regression, 
where the outcome is the dependent variable. The causal effect estimate, then, is the 
coefficient $\hat \beta_1$ obtained from the second regression analysis (Equation \@ref(eq:lm2)). Note also that the random errors in Equation \@ref(eq:lm2) are assumed independent of those in Equation \@ref(eq:lm1). Like the $\epsilon_i$ in Equation \@ref(eq:lm1), the $\tau_i$ are assumed independent and identically distributed normal random errors.

@@

@@

@@

Likelihood-based MR methods, unlike two-stage methods, provide maximum likelihood 
estimates with their many desirable properties [@fisher1925theory]. Limited information maximum likelihood from econometrics 
is the earliest approach for likelihood-based inference in MR [@davidson1993estimation]. CoMM and 
PMR-Egger are two recent likelihood-based methods [@yang_comm_2019;@yuan_testing_2020]. We next discuss CoMM and CoMM-S2.

@@

It is worthwhile to consider ways in which the standard MR assumptions may be violated. 
Let's first consider the exclusion restriction assumption (assumption 2 above). It states that the genetic marker 
must affect the outcome only through the exposure. The causal diagram in Figure \@ref(fig:dag1) illustrates 
a scenario that violates this assumption when the variable X is used as the exposure [@vanderweele_methodological_2014]. Bias in causal effect estimation would result. However, 
if variable W is measured and treated as the exposure, then an unbiased estimate of causal effect is 
possible, since W captures the two pathways, W -> Y and W -> X -> Y, through which the genetic marker G 
affects the outcome Y [@vanderweele_methodological_2014;@textor2011dagitty;@textor2016robust].

@@

Sensitivity analysis is recommended and widely used in MR studies because of the inability to verify 
the three MR assumptions with observational data. The goal of a sensitivity analysis is to gain insight 
into how the results might differ if the assumptions be violated.
Because not all confounding 
variables are known or measured, assumptions 2 and 3 are not fully verifiable 
in MR studies. Recognition of this fact has inspired the development of 
sensitivity analysis tools. While we can't assess whether a SNP instrument is 
associated with every confounder of the exposure-outcome association, it is 
possible to examine the associations of the SNP instrument with the measured 
covariates. While absence of such associations doesn't guarantee satisfaction 
of the assumption, presence of SNP-covariate associations must be investigated
carefully, as they may constitute assumption violations [@burgess_sensitivity_2017].

@@

Recent research on the omnigenic hypothesis, which posits 
that every SNP's effect on a trait is nonzero, has informed
MR methods development [@Boyle2017]. @wang_mendelian_nodate 
recognized the limitations of previous MR methods in the 
context of the omnigenic hypothesis. In response, @wang_mendelian_nodate 
developed a MR method that uses all genome-wide SNPs as instruments. 
Using GWAS summary statistics as inputs, their method relies on a composite likelihood framework for scalable computation and allows for horizontal pleiotropy. @wang_mendelian_nodate used extensive 
simulations, including those with model misspecifications, to characterize their method's statistical power and robustness. Finally, they applied the new method to 
identify multiple complex traits that affect coronary artery disease and asthma. These causal relationships highlight the important roles of plasma lipids, blood pressure, and the immune system in CAD susceptibility and those of obesity and the immune system in asthma development.
